---
title: "Sampling Distributions"
author: "Cheng Peng"
date: "West Chester University"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: yes
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    theme: lumen
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 3
    fig_height: 3
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
editor_options: 
  chunk_output_type: inline
---

```{css, echo = FALSE}
#TOC::before {
  content: "Table of Contents";
  font-weight: bold;
  font-size: 1.2em;
  display: block;
  color: navy;
  margin-bottom: 10px;
}


div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
}

h1.title {    /* level 1 header of title  */
  font-size: 22px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 15px;
  font-weight: bold;
  font-family: system-ui;
  color: navy;
  text-align: center;
}

h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Gill Sans", sans-serif;
  color: DarkBlue;
  text-align: center;
}

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 20px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}

h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 16px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 14px;
  font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Add dots after numbered headers */
.header-section-number::after {
  content: ".";

body { background-color:white; }

.highlightme { background-color:yellow; }

p { background-color:white; }

}
```

```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("pander")) {
   install.packages("pander")
   library(pander)
}
if (!require("ggplot2")) {
  install.packages("ggplot2")
  library(ggplot2)
}
if (!require("tidyverse")) {
  install.packages("tidyverse")
  library(tidyverse)
}

if (!require("plotly")) {
  install.packages("plotly")
  library(plotly)
}

## library(leaps)
knitr::opts_chunk$set(echo = TRUE,       # include code chunk in the output file
                      warning = FALSE,   # sometimes, you code may produce warning messages,
                                         # you can choose to include the warning messages in
                                         # the output file. 
                      results = TRUE,    # you can also decide whether to include the output
                                         # in the output file.
                      message = FALSE,
                      comment = NA
                      )  
```

\


# Introduction

Sampling distributions form the cornerstone of statistical inference. They describe the probability distribution of a **sample statistic** calculated from random samples. This note explores both exact (finite-sample) and asymptotic (large-sample) distributions for key statistics including sample means, proportions, and related test statistics.


# Sampling Distribution of the Sample Mean

When the population is normal, by the property of normal distribution, the sum of the iid random variables are **exactly** normally distributed. If the population is not a normal distribution, using the Central Limit Theorem (CLT), the sum of the iid random variables is **asymptotically** normally distributed.


## Exact Distribution

For a random sample $X_1, X_2, \ldots, X_n$ from a normal population $N(\mu, \sigma^2)$, the sample mean has an exact normal distribution:

$$
\bar{X} \to N\left(\mu,  \frac{\sigma}{\sqrt{n}}\right)
$$

The standardized version is:

$$
Z = \frac{\bar{X}-\mu}{\sigma/\sqrt{n}} \to N(0, 1)
$$

**Example**:

```{r}
set.seed(123)
n <- 10
mu <- 5
sigma <- 2

n.samples <- 10000
sample.means <- replicate(n.samples, mean(rnorm(n, mu, sigma)))

# Create theoretical curve data
x.vals <- seq(mu - 3*sigma/sqrt(n), mu + 3*sigma/sqrt(n), length.out = 100)
theory.density <- dnorm(x.vals, mean = mu, sd = sigma/sqrt(n))
theory.df <- data.frame(x = x.vals, density = theory.density)

xbar.plt <- ggplot(data.frame(mean = sample.means), aes(x = mean)) +
  geom_histogram(aes(y = ..density..), bins = 50, alpha = 0.7, fill = "gray") +
  geom_line(data = theory.df, aes(x = x, y = density), 
            color = "red", linewidth = 1) +
  labs(title = "Exact Sampling Distribution of Sample Mean \nNormal Population (n = 10)",
       x = "Sample Mean", y = "Density") +
   theme(plot.title = element_text(hjust = 0.5),
        plot.margin = margin(t = 35, r = 20, b = 30, l = 30, unit = "pt"))

ggplotly(xbar.plt)

```


## Asymptotic Distribution (Central Limit Theorem)

For any population with finite mean $\mu$ and variance $\sigma^2$, as $n \to \infty$:



$$
Z = \frac{\bar{X}-\mu}{\sigma/\sqrt{n}} \to_{\text{approx}} N(0, 1)
$$


**Example** Consider exponential population:

```{r}
set.seed(123)
n.large <- 50
lambda <- 1/5  # Mean = 5

# Generate multiple samples from exponential distribution
n.samples <- 10000
exp.means <- replicate(n.samples, mean(rexp(n.large, rate = lambda)))

# Compare with normal approximation
theoretical.mean <- 1/lambda  # 5
theoretical.sd <- (1/lambda)/sqrt(n.large)  # 5/sqrt(50)

theory.density <- dnorm(x.vals, mean = theoretical.mean, sd = theoretical.sd)
theory.df <- data.frame(x = x.vals, density = theory.density)

# Option 1: Use only stat_function for theoretical curve (Recommended)
gg.clt <- ggplot(data.frame(mean = exp.means), aes(x = mean)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, alpha = 0.7, fill = "lightgreen") +
  geom_line(data = theory.df, aes(x = x, y = density), 
            color = "red", linewidth = 1) +
  labs(title = "Asymptotic Sampling Distribution of Sample Mean \nExponential Population (n = 50)",
       x = "Sample Mean", y = "Density") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.margin = margin(t = 35, r = 20, b = 30, l = 30, unit = "pt"))
#gg.clt
ggplotly(gg.clt)
```






# Student's t-Distribution

When population variance $\sigma^2$ is unknown and estimated by sample variance $S^2$:

$$
T = \frac{\bar{X}-\mu}{S/\sqrt{n}} \to  t_{n-1}
$$

where $S^2 = \frac{1}{n-1}\sum_{i=1}^n (X_i - \bar{X})^2$



**Example**:

```{r}
set.seed(123)
n <- 10
mu <- 5
sigma <- 2

# Generate t-statistics
n.samples <- 10000
t.stats <- numeric(n.samples)

for(i in 1:n.samples) {
  sample.data <- rnorm(n, mu, sigma)
  x.bar <- mean(sample.data)
  s <- sd(sample.data)
  t.stats[i] <- (x.bar - mu) / (s/sqrt(n))
}

# Compare with theoretical t-distribution
x.vals <- seq(-4, 4, length.out = 200)
theoretical.t <- dt(x.vals, df = n-1)
theoretical.normal <- dnorm(x.vals)

comparison.df <- data.frame(
  x = rep(x.vals, 2),
  density = c(theoretical.t, theoretical.normal),
  distribution = rep(c("t(9)", "N(0,1)"), each = length(x.vals))
)

t.plt <- ggplot(comparison.df, aes(x = x, y = density, color = distribution)) +
  geom_line(size = 1) +
  labs(title = "t-Distribution vs Normal Distribution",
       x = "Value", y = "Density") +
    theme(plot.title = element_text(hjust = 0.5),
        plot.margin = margin(t = 35, r = 20, b = 30, l = 30, unit = "pt")) +
   scale_color_manual(values = c("red", "blue"))
ggplotly(t.plt)
```



# Sampling Distribution of Sample Proportion

3.1 Exact Distribution

For a binomial population with success probability $p$, the sample proportion $\hat{p} = X/n$ where $X \sim Binomial(n,p)$.

The exact distribution is simply the probability mass function of a binomial distribution with n trials and success probability $p$:

$$
P(\hat{p} =k/n)=P(X = k)=\frac{n!}{k!(n-k)!} p^k (1−p)^{n-k}, \ \ k = 0, 1, 2, \cdots, n.
$$ 


## Asymptotic Distribution

By Central Limit Theorem, as $n \to \infty$:

$$
\hat{p} \to N\left(p, \sqrt{\frac{p(1-p)}{n}} \right)
$$

**Example**:

```{r}
set.seed(123)
n <- 100
p <- 0.3

# Generate sample proportions
n.samples <- 10000
sample.props <- replicate(n.samples, rbinom(1, n, p)/n)

# Compare with normal approximation
theoretical.mean <- p
theoretical.sd <- sqrt(p*(1-p)/n)

x.vals <- seq(0,0.6, length=100)
theory.density <- dnorm(x.vals, mean = theoretical.mean, sd = theoretical.sd)
theory.df <- data.frame(x = x.vals, density = theory.density)

binom.plt <- ggplot(data.frame(prop = sample.props), aes(x = prop)) +
  geom_histogram(aes(y = ..density..), bins = 30, alpha = 0.7, fill = "skyblue") +
  geom_line(data = theory.df, aes(x = x, y = density), 
            color = "red", linewidth = 1) +
  #stat_function(fun = dnorm, 
  #              args = list(mean = theoretical_mean, sd = theoretical_sd),
  #              color = "red", size = 1) +
  labs(title = "Sampling Distribution of Sample Proportion",
       subtitle = "p = 0.3, n = 100",
       x = "Sample Proportion", y = "Density") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.margin = margin(t = 35, r = 20, b = 30, l = 30, unit = "pt"))
ggplotly(binom.plt)
```


# Chi-Square Distribution

For $Z_1, Z_2, \ldots, Z_k \stackrel{iid}{\sim} N(0,1)$, using moment generating function, we can show that

$$
Q=\sum_{i=1}^k Z_i^2 \to \chi_k^2.
$$
 

For sample variance from normal population:

$$
\frac{(n-1)S^2}{\sigma^2} \to \chi_{n-1}^2
$$
**Proof**: We prove this in several steps:

We show that for $X_1, \dots, X_n \iid N(\mu, \sigma^2)$, with

$$
S^2 = \frac{1}{n-1} \sum_{i=1}^n (X_i - \bar{X})^2, \quad \bar{X} = \frac{1}{n} \sum_{i=1}^n X_i,
$$

we have

$$
\frac{(n-1)S^2}{\sigma^2} \sim \chi_{n-1}^2.
$$

**Step 1: Standardize and define notation**

Let $Z_i = \frac{X_i - \mu}{\sigma} \sim N(0,1)$, i.i.d. Then

$$
\bar{Z} = \frac{1}{n} \sum_{i=1}^n Z_i = \frac{\bar{X} - \mu}{\sigma}.
$$

We can write:

$$
\sum_{i=1}^n (X_i - \bar{X})^2 = \sigma^2 \sum_{i=1}^n (Z_i - \bar{Z})^2.
$$

So

$$
\frac{(n-1)S^2}{\sigma^2} = \frac{\sum_{i=1}^n (X_i - \bar{X})^2}{\sigma^2} = \sum_{i=1}^n (Z_i - \bar{Z})^2.
$$

**Step 2: Orthogonal transformation**

Let \( \mathbf{Z} = (Z_1, \dots, Z_n)^T \). Choose an \( n \times n \) orthogonal matrix \( Q \) whose first row is \( \left( \frac{1}{\sqrt{n}}, \dots, \frac{1}{\sqrt{n}} \right) \). Define

$$
\mathbf{Y} = Q \mathbf{Z}.
$$

Then:

* $Y_1 = \frac{1}{\sqrt{n}} \sum_{i=1}^n Z_i = \sqrt{n} \, \bar{Z}$.
* Since $Q$ is orthogonal and $\mathbf{Z} \sim N(0, I_n)$, we have $\mathbf{Y} \sim N(0, I_n)$ as well, so $Y_1, \dots, Y_n$ are i.i.d.\ $N(0,1)$.


**Step 3: Express sum of squares in terms of \( Y_j \)**

Orthogonality implies:

$$
\sum_{i=1}^n Z_i^2 = \sum_{j=1}^n Y_j^2.
$$

Also,

$$
\sum_{i=1}^n (Z_i - \bar{Z})^2 = \sum_{i=1}^n Z_i^2 - n \bar{Z}^2.
$$

But $n \bar{Z}^2 = Y_1^2$, so

$$
\sum_{i=1}^n (Z_i - \bar{Z})^2 = \sum_{j=1}^n Y_j^2 - Y_1^2 = \sum_{j=2}^n Y_j^2.
$$

**Step 4: Distribution**

Since $Y_2, \dots, Y_n$ are i.i.d.\ $N(0,1)$, we have

$$
\sum_{j=2}^n Y_j^2 \sim \chi_{n-1}^2.
$$

Thus

$$
\frac{(n-1)S^2}{\sigma^2} = \sum_{i=1}^n (Z_i - \bar{Z})^2 = \sum_{j=2}^n Y_j^2 \sim \chi_{n-1}^2.
$$

**Step 5: Independence from \( \bar{X} \)**

Since $Y_1 = \sqrt{n} \bar{Z}$ is independent of $Y_2, \dots, Y_n$, it follows that $\bar{X}$ is independent of $S^2$. That is,

$$
\boxed{\frac{(n-1)S^2}{\sigma^2} \to \chi_{n-1}^2}
$$

**Example**: The $\chi^2$ distribution is derived from the standard normal distribution. We simulate standard normal random numbers and then transform them into $\chi^2$ random variables based on the derivations above. A histogram will be plotted and overlaid with the theoretical $\chi^2$ density curve.

```{r}
set.seed(123)
n <- 10
sigma <- 2

# Generate chi-square statistics
n.samples <- 10000
chisq.stats <- numeric(n.samples)

for(i in 1:n.samples) {
  sample.data <- rnorm(n, 0, sigma)
  chisq.stats[i] <- sum((sample.data/sigma)^2)
}

# Compare with theoretical chi-square
x.vals <- seq(0, 30, length.out = 200)
theoretical.chisq <- dchisq(x.vals, df = n)
theory.df <- data.frame(x = x.vals, density = theoretical.chisq)

chi.plt <- ggplot(data.frame(x = chisq.stats), aes(x = x)) +
  geom_histogram(aes(y = ..density..), bins = 50, alpha = 0.7, fill = "steelblue") +
  geom_line(data = theory.df, aes(x = x, y = density), 
            color = "red", linewidth = 1) +
  #stat_function(fun = dchisq, args = list(df = n), color = "red", size = 1) +
  labs(title = "Chi-Square Distribution",
       subtitle = "Sum of squared standard normals",
       x = "Value", y = "Density") +
   theme(plot.title = element_text(hjust = 0.5),
        plot.margin = margin(t = 35, r = 20, b = 30, l = 30, unit = "pt"))
ggplotly(chi.plt)
```

# F-Distribution

For two independent chi-square random variables:

$$
F = \frac{U_1/d_1}{U_2/d_2} \to F_{d_1, d_2}
$$
 
where $U_1 \sim \chi^2_{d_1}$ and $U_2 \sim \chi^2_{d_2}$.



F distribution is used for comparing variances: $\frac{S_1^2/\sigma_1^2}{S_2^2/\sigma_2^2} \sim F_{n_1-1,n_2-1}$. For example, if we test

$$
H_0: \ \sigma_1 = \sigma_2 \ \ \ v.s. \ \ \ H_a:  \sigma_1 \ne \sigma_2
$$
The test statistic

$$
TS = \frac{S_1^2}{S_2^2} \to F_{n_1-1, n_2-1}
$$


**Example**: The F distribution is directly defined based on two independent $\chi^2$ distributions, which are themselves derived from standard normal distributions. Therefore, we could generate data from normal distributions and then transform them into F random variables. To keep the process simple, we generate data directly from $\chi^2$ distributions.

```{r}
set.seed(123)
df1 <- 10
df2 <- 15

# Generate F statistics
n.samples <- 10000
f.stats <- numeric(n.samples)

for(i in 1:n.samples) {
  u1 <- rchisq(1, df1)
  u2 <- rchisq(1, df2)
  f.stats[i] <- (u1/df1) / (u2/df2)
}

# Compare with theoretical F-distribution
x.vals <- seq(0, 5, length.out = 200)
theoretical.f <- df(x.vals, df1, df2)
theory.df <- data.frame(x = x.vals, density = theoretical.f)




f.plt <- ggplot(data.frame(x = f.stats), aes(x = x)) +
  geom_histogram(aes(y = ..density..), bins = 50, alpha = 0.7, fill = "purple3") +
  geom_line(data = theory.df, aes(x = x, y = density), 
            color = "red", linewidth = 1) +
  coord_cartesian(xlim = c(0, 5)) +
  labs(title = paste("F-Distribution \n F(", df1, ",", df2, ")", sep = ""),
       x = "Value", y = "Density") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.margin = margin(t = 35, r = 20, b = 30, l = 30, unit = "pt"))
ggplotly(f.plt)
```



# Summary of Key Relationships

|Statistic	| Exact Distribution |	Asymptotic Distribution |	Conditions |
|:----------|:--------------|:--------------------|:-------------|
| $\bar{X}$	| $N(\mu, \sigma^2/n)$| 	$N(\mu, \sigma^2/n)$| 	Normal population or large n| 
| $\frac{\bar{X}-\mu}{S/\sqrt{n}}$| 	$t_{n-1}$	| $N(0,1)$| 	Normal population| 
| $\hat{p}$	| $Binomial(n,p)/n$	| $N(p, p(1-p)/n)$| 	$np, n(1-p) \geq 5$| 
| $\frac{(n-1)S^2}{\sigma^2}$	| $\chi^2_{n-1}$ |-	| Normal population| 
| $\frac{S_1^2/\sigma_1^2}{S_2^2/\sigma_2^2}$| $F_{n_1-1,n_2-1}$| 	-	| Normal populations| 


**Conclusion**

* Understanding sampling distributions is fundamental to statistical inference:

* Exact distributions provide precise results when assumptions are met

* Asymptotic distributions offer approximations for large samples

* The choice between exact and asymptotic methods depends on sample size, distributional assumptions, and the specific parameter being estimated

* Modern computing allows for empirical verification of these theoretical results


These distributions form the theoretical foundation for hypothesis testing, confidence intervals, and many other statistical procedures.





