---
title: "Bootstrap Testing Hypothesis"
author: "Cheng Peng"
date: "West Chester University"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: yes
    toc_collapsed: yes
    code_folding: show
    code_download: yes
    smooth_scroll: yes
    theme: lumen
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 3
    fig_height: 3
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
editor_options: 
  chunk_output_type: inline
---

```{css, echo = FALSE}
#TOC::before {
  content: "Table of Contents";
  font-weight: bold;
  font-size: 1.2em;
  display: block;
  color: navy;
  margin-bottom: 10px;
}


div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
}

h1.title {    /* level 1 header of title  */
  font-size: 22px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 15px;
  font-weight: bold;
  font-family: system-ui;
  color: navy;
  text-align: center;
}

h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Gill Sans", sans-serif;
  color: DarkBlue;
  text-align: center;
}

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 20px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}

h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 16px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 14px;
  font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Add dots after numbered headers */
.header-section-number::after {
  content: ".";

body { background-color:white; }

.highlightme { background-color:yellow; }

p { background-color:white; }

}
```

```{html, echo=FALSE}
<button id="toggleTOCBtn" class="btn btn-default" style="position: fixed; bottom: 20px; right: 20px; z-index: 1000;">
  Toggle TOC
</button>

<script>
document.getElementById("toggleTOCBtn").onclick = function() {
  var toc = document.querySelector(".list-group, #TOC, .tocify");
  if (toc) {
    if (toc.style.display === "none") {
      toc.style.display = "";
    } else {
      toc.style.display = "none";
    }
  }
};
</script>

```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("pander")) {
   install.packages("pander")
   library(pander)
}
if (!require("ggplot2")) {
  install.packages("ggplot2")
  library(ggplot2)
}
if (!require("tidyverse")) {
  install.packages("tidyverse")
  library(tidyverse)
}

if (!require("plotly")) {
  install.packages("plotly")
  library(plotly)
}
if (!require("fitdistrplus")) {
  install.packages("fitdistrplus")
  library(fitdistrplus)
}
## library(fitdistrplus)
knitr::opts_chunk$set(echo = TRUE,       # include code chunk in the output file
                      warning = FALSE,   # sometimes, you code may produce warning messages,
                                         # you can choose to include the warning messages in
                                         # the output file. 
                      results = TRUE,    # you can also decide whether to include the output
                                         # in the output file.
                      message = FALSE,
                      comment = NA
                      )  
```

\

# Introduction


This module presents comprehensive methods for bootstrap hypothesis testing. For each test (one-sample, two-sample, and alternative Wald, Score, and likelihood ratio tests), we state the mathematical algorithm/steps and provide pseudo-code before presenting numerical examples. All implementations are provided in R with complete code examples.

Bootstrap hypothesis testing involves resampling data to **approximate the sampling distribution** of a test statistic under **the null hypothesis**. The general procedure consists of:

* Compute the test statistic from the original data: $T_{obs}$
* Transform/resample data to satisfy the <font color = "red">**null hypothesis $H_0$**</font>
* Generate $B$ bootstrap samples and compute test statistics $T_b^*$
* Calculate p-value (for a two-tailed test) as: 

$$
p = \frac{\#\{|T_b^*| \geq |T_{obs}|\} + 1}{B + 1}
$$

The bootstrap provides a nonparametric approach to inference that doesn't rely heavily on distributional assumptions (i.e., the sampling distribution of the test statistic).

# One-Sample Bootstrap Tests

## One-Sample Mean Test (Parametric Bootstrap)

**Mathematical Algorithm**

Let $X = \{x_1, x_2, ..., x_n\}$ be the observed data.


* **Null hypothesis**: $H_0: \mu = \mu_0$

* **Test statistic**:  for $\bar{x} = \frac{1}{n}\sum_{i=1}^n x_i$ and $s = \sqrt{\frac{1}{n-1}\sum_{i=1}^n (x_i - \bar{x})^2}$, the test statistic is defined to be
$$
t_{obs} = \frac{\bar{x} - \mu_0}{s/\sqrt{n}}
$$

* **Center data under $H_0$**: $x_i^\prime = x_i - \bar{x} + \mu_0$.  For $b = 1$ to $B$}:
  + Sample $X_b^* = \{x_1^*, ..., x_n^*\}$ from $X'$ with replacement
  
$$\bar{x}_b^* = \frac{1}{n}\sum_{i=1}^n x_i^*$$
$$s_b^* = \sqrt{\frac{1}{n-1}\sum_{i=1}^n (x_i^* - \bar{x}_b^*)^2}$$
$$t_b^* = \frac{\bar{x}_b^* - \mu_0}{s_b^*/\sqrt{n}}$$


* **Calculate p-value(for two-sided test):
$$
p = \frac{\#\{|t_b^*| \geq |t_{obs}|\} + 1}{B + 1}
$$


**Pseudo-code**

\begin{algorithm}
\caption{One-Sample Mean Bootstrap Test}
\begin{algorithmic}[1]
\Function{one\_sample\_mean\_test}{data, $\mu_0$, $B=9999$}
    \State $n \gets \text{length}(data)$
    \State $t_{obs} \gets (\text{mean}(data) - \mu_0) / (\text{sd}(data)/\sqrt{n})$
    
    \Comment{Center data under $H_0$}
    \State $centered\_data \gets data - \text{mean}(data) + \mu_0$
    
    \State $t^* \gets \text{numeric}(B)$
    \For{$b = 1$ to $B$}
        \State $bootstrap\_sample \gets \text{sample}(centered\_data, n, \text{replace}=TRUE)$
        \State $t^*[b] \gets (\text{mean}(bootstrap\_sample) - \mu_0) / (\text{sd}(bootstrap\_sample)/\sqrt{n})$
    \EndFor
    
    \State $p\_value \gets (\text{sum}(|t^*| \geq |t_{obs}|) + 1) / (B + 1)$
    \State \Return $p\_value, t_{obs}, t^*$
\EndFunction
\end{algorithmic}
\end{algorithm}


**R Implementation**

```{r}
# One-sample mean test using parametric bootstrap
one_sample_bootstrap_test <- function(data, mu0, B = 9999) {
    n <- length(data)
    t_obs <- (mean(data) - mu0) / (sd(data) / sqrt(n))
    
    # Center data under H0: μ = μ0
    centered_data <- data - mean(data) + mu0
    
    t_star <- numeric(B)
    for (b in 1:B) {
        bs_sample <- sample(centered_data, n, replace = TRUE)
        t_star[b] <- (mean(bs_sample) - mu0) / (sd(bs_sample) / sqrt(n))
    }
    
    p_value <- (sum(abs(t_star) >= abs(t_obs)) + 1) / (B + 1)
    
    return(list(
        p_value = p_value,
        t_obs = t_obs,
        bootstrap_stats = t_star,
        mean_obs = mean(data),
        mu0 = mu0
    ))
}

# Example
set.seed(123)
data <- rnorm(30, mean = 105, sd = 15)
mu0 <- 100

result <- one_sample_bootstrap_test(data, mu0, B = 1999)
```


## One-Sample Mean Test (Percentile Method)

**Mathematical Algorithm**

* Compute sample mean: $\bar{x} = \frac{1}{n}\sum_{i=1}^n x_i$, For $b = 1$ to $B$:
  + Sample $X_b^*$ from $X$ with replacement
  + $\bar{x}_b^* = \frac{1}{n}\sum_{i=1}^n x_i^*$

* Construct $(1-\alpha)\%$ confidence interval from percentiles. For $\bar{x}_{(q)}$ is the $q$-th quantile of $\{\bar{x}_b^*\}$
$$
CI = [\bar{x}_{(\alpha/2)}, \bar{x}_{(1-\alpha/2)}]
$$

* Decision rule: Reject $H_0: \mu = \mu_0$ if $\mu_0 \notin CI$
* P-value approximation:
$$
p \approx 2 \times \min\left[P(\bar{x}^* \leq \mu_0), P(\bar{x}^* \geq \mu_0)\right]
$$



# Two-Sample Bootstrap Tests

**Mathematical Algorithm**

Let $X = \{x_1, ..., x_m\}$ and $Y = \{y_1, ..., y_n\}$ be two independent samples.

* **Null hypothesis**: $H_0: \mu_x = \mu_y$
* **Pooled standard deviation** (under $H_0$):
$$
s_p = \sqrt{\frac{(m-1)s_x^2 + (n-1)s_y^2}{m+n-2}}
$$

* Test statistic}:
$$
t_{obs} = \frac{\bar{x} - \bar{y}}{s_p\sqrt{\frac{1}{m} + \frac{1}{n}}}
$$

* **Center and pool data under $H_0$}**:
$$z_i = \begin{cases}
x_i - \bar{x} & \text{for } i = 1,...,m \\
y_i - \bar{y} & \text{for } i = 1,...,n
\end{cases}$$

* For $b = 1$ to $B$:
  +  $X_b^* = \{z_1^*, ..., z_m^*\}$ from pooled data with replacement
  +  Sample $Y_b^* = \{z_{m+1}^*, ..., z_{m+n}^*\}$ from pooled data with replacement
  +  Compute: for  $s_{p,b}^*$ is the pooled SD of $X_b^*$ and $Y_b^*$
$$
t_b^* = \frac{\bar{x}_b^* - \bar{y}_b^*}{s_{p,b}^*\sqrt{\frac{1}{m} + \frac{1}{n}}}
$$

* **Calculate p-value**:
$$
p = \frac{\#\{|t_b^*| \geq |t_{obs}|\} + 1}{B + 1}
$$



# Likelihood-based Bootstrap Tests

This section presents Bootstrap version of the three likelihood-based $\chi^2$ tests.

## Bootstrap Wald Test

**Mathematical Formulation**

For a parameter $\theta$ with null hypothesis $H_0: \theta = \theta_0$:

* **Maximum likelihood estimator**: $\hat{\theta}$
* **Estimated variance** $\widehat{Var}(\hat{\theta})$
* **Wald statistic**
$$
W = \frac{(\hat{\theta} - \theta_0)^2}{\widehat{Var}(\hat{\theta})}
$$
Under $H_0$, $W \sim \chi^2_1$ asymptotically



**Bootstrap Algorithm**

*  Compute $\hat{\theta}$ and $W_{obs}$ from original data
*  For $b = 1$ to $B$:
  +  Generate bootstrap sample $X_b^*$
  +  Compute $\hat{\theta}_b^*$ from $X_b^*$
  +  Estimate $\widehat{Var}_b^*(\hat{\theta}_b^*)$ using nested bootstrap
  +  Compute:
$$
W_b^* = \frac{(\hat{\theta}_b^* - \hat{\theta})^2}{\widehat{Var}_b^*(\hat{\theta}_b^*)}
$$

* Calculate p-value:
$$
p = \frac{\#\{W_b^* \geq W_{obs}\} + 1}{B + 1}
$$


**R Implementation**

```{r}
# Bootstrap Wald test for mean
bootstrap_wald_test <- function(data, theta0, B = 999, inner_B = 100) {
    n <- length(data)
    theta_hat <- mean(data)
    
    # Bootstrap for variance estimation
    boot_theta <- numeric(B)
    for (b in 1:B) {
        bs_sample <- sample(data, n, replace = TRUE)
        boot_theta[b] <- mean(bs_sample)
    }
    
    # Wald statistic
    var_theta <- var(boot_theta)
    W_obs <- (theta_hat - theta0)^2 / var_theta
    
    # Bootstrap distribution of Wald statistic
    W_star <- numeric(B)
    for (b in 1:B) {
        bs_sample <- sample(data, n, replace = TRUE)
        theta_star <- mean(bs_sample)
        
        # Inner bootstrap for variance estimation
        inner_boot <- numeric(inner_B)
        for (j in 1:inner_B) {
            inner_sample <- sample(bs_sample, n, replace = TRUE)
            inner_boot[j] <- mean(inner_sample)
        }
        var_star <- var(inner_boot)
        
        W_star[b] <- (theta_star - theta_hat)^2 / var_star
    }
    
    p_value <- (sum(W_star >= W_obs) + 1) / (B + 1)
    
    return(list(
        W_obs = W_obs,
        p_value = p_value,
        theta_hat = theta_hat,
        theta0 = theta0
    ))
}
```


## Bootstrap Score Test 

**Mathematical Formulation**

For a parameter $\theta$:

* **Score function**: $U(\theta) = \frac{\partial \ell(\theta)}{\partial \theta}$
where $\ell(\theta)$ is the log-likelihood
* **Fisher information**: $I(\theta) = -E\left[\frac{\partial^2 \ell(\theta)}{\partial \theta^2}\right]$
* **Score statistic}**: Under $H_0: \theta = \theta_0$, the following score test statistic $S \sim \chi^2_1$ asymptotically
$$
S = \frac{U(\theta_0)^2}{I(\theta_0)}
$$


**Bootstrap Algorithm**

* Compute $U(\theta_0)$ and $I(\theta_0)$ under $H_0$
* Compute $S_{obs}$
* For $b = 1$ to $B$:
  + Generate data under $H_0$: $X_b^* \sim f(x|\theta_0)$
  + Compute $U_b(\theta_0)$ from $X_b^*$
  + Compute:
$$
S_b^* = \frac{U_b(\theta_0)^2}{I(\theta_0)}
$$

* Calculate p-value:
$$
p = \frac{\#\{S_b^* \geq S_{obs}\} + 1}{B + 1}
$$
 

## Bootstrap Likelihood Ratio Test 

**Mathematical Formulation**

* **Likelihood function**: $L(\theta) = \prod_{i=1}^n f(x_i|\theta)$
* **Maximum likelihood estimator**: $\hat{\theta} = \arg\max_\theta L(\theta)$
* **Likelihood ratio**:
$$
\Lambda = \frac{L(\theta_0)}{L(\hat{\theta})}
$$
* **Likelihood ratio statistic**: Under $H_0: \theta = \theta_0$, $LR = -2\log\Lambda \sim \chi^2_1$ asymptotically distributed as $\chi^2_{d.f}$
$$
-2\log\Lambda = -2[\log L(\theta_0) - \log L(\hat{\theta})]
$$



**Bootstrap Algorithm**

* Compute $-2\log\Lambda_{obs}$ from original data
* For $b = 1$ to $B$:
  +  Generate data under $H_0$: $X_b^* \sim f(x|\theta_0)$
  + Compute MLE $\hat{\theta}_b^*$ from $X_b^*$
  + Compute $-2\log\Lambda_b^*$

*  Calculate p-value:
$$
p = \frac{\#\{-2\log\Lambda_b^* \geq -2\log\Lambda_{obs}\} + 1}{B + 1}
$$


**R Implementation**

```{r}
# Bootstrap Likelihood Ratio test for normal mean
bootstrap_lrt_test <- function(data, mu0, B = 9999) {
    n <- length(data)
    
    # MLEs (assuming unknown variance)
    mu_hat <- mean(data)
    sigma_hat <- sd(data) * sqrt((n - 1) / n)  # MLE of σ
    
    # Using same sigma estimate for both for fair comparison
    sigma0 <- sd(data) * sqrt((n - 1) / n)
    
    # Log-likelihoods
    logL_hat <- sum(dnorm(data, mean = mu_hat, sd = sigma_hat, log = TRUE))
    logL0 <- sum(dnorm(data, mean = mu0, sd = sigma0, log = TRUE))
    
    # LR statistic
    LR_obs <- -2 * (logL0 - logL_hat)
    
    # Bootstrap distribution
    LR_star <- numeric(B)
    for (b in 1:B) {
        # Generate data under H0: N(μ0, σ0)
        bs_sample <- rnorm(n, mean = mu0, sd = sigma0)
        
        # MLE from bootstrap sample
        mu_star <- mean(bs_sample)
        sigma_star <- sd(bs_sample) * sqrt((n - 1) / n)
        
        # Likelihoods
        logL_star <- sum(dnorm(bs_sample, mean = mu_star, sd = sigma_star, log = TRUE))
        logL_star0 <- sum(dnorm(bs_sample, mean = mu0, sd = sigma0, log = TRUE))
        
        LR_star[b] <- -2 * (logL_star0 - logL_star)
    }
    
    p_value <- (sum(LR_star >= LR_obs) + 1) / (B + 1)
    
    return(list(
        LR_obs = LR_obs,
        p_value = p_value,
        mu_hat = mu_hat,
        mu0 = mu0
    ))
}
```


## Non-Permutation Test with Bootstrap

**Mathematical Algorithm**

* **Combine groups**: $Z = \{x_1, ..., x_m, y_1, ..., y_n\}$
* **Observed statistic**: $T_{obs} = \bar{x} - \bar{y}$
* For $b = 1$ to $B$}:
  + Randomly permute $Z$ to get $Z_b^*$
  + Assign first $m$ elements to $X_b^*$, remaining $n$ to $Y_b^*$
  + Compute $T_b^* = \bar{x}_b^* - \bar{y}_b^*$

* **Calculate p-value**:
$$
p = \frac{\#\{|T_b^*| \geq |T_{obs}|\} + 1}{B + 1}
$$



**Choosing Number of Bootstrap Samples**

The choice of $B$ (number of bootstrap samples) affects the precision of p-value estimation. Recommended values:

* For p-values near 0.05: $B \geq 999$
* For publication quality: $B \geq 9999$
* For confidence intervals: $B \geq 1000$ for 95\% CI, more for higher confidence levels
 

The standard error of a bootstrap p-value is approximately:
$$
\text{SE}(p) = \sqrt{\frac{p(1-p)}{B}}
$$

For $p = 0.05$ and $B = 1000$, $\text{SE} \approx 0.0069$, giving about 1.4 percentage points of uncertainty.

# Conclusion

This document has presented comprehensive methods for bootstrap hypothesis testing in R, covering one-sample tests, two-sample tests, and alternative test statistics (Wald, Score, and likelihood ratio tests). Each method includes mathematical formulation, algorithmic steps, pseudo-code, and practical R implementations. Bootstrap methods provide robust alternatives to classical tests, especially when distributional assumptions are violated or sample sizes are small.


